# ECE608 Assignment 3
Natalia Hoyos Velasquez 
UW ID

# Data set
The Assignment aims to use the Board game rankings from boardgamegeek.com dataset to gain experience on the use of the R environment and on working with tabular data. This dataset is publicly available and was originally obtained by scrapping information from boardgamegeek.com.
The file board_games.csv was taken from [1] where the data set can also be consulted. The csv file was used as the data source and the documentation was used to understand the dataset's contents and to identify the column labels.

# Tools 
R 3.6.0 was used for this assignment, in Ubuntu 19.04 using RStudio. The tidyverse, ez, e1071, car, and emmeans libraries for R were loaded. Tidyverse enables the use of ggplot2, dplyr, tidyr, and readr packages for diverse objectives including importing, manipulating and plotting data. Ez is imported to carry out the ANOVA test and e1071 is useful for its miscellaneous statistical functions.

```{r}
library("gridExtra")
library("grid")
library(tidyverse)    #For data manipulation
library(ez)           #For ANOVA
library(e1071)        #Misc statistical functions
library(car)          #For qqplot
```

# Data Set

```{r}
#Read and clean data
df <- read.csv('D:/Google Drive/AA UW/2019-05/ECE 608 Quantitative Methods in BME/Assignments/Assignment 3/ECE608_A3_BloodFlow.csv')

head(df)

str(df)
writeLines(paste("\nMissing values:", sum(is.na(df))))
```

The structure function (str( ) ) gives a glance at the data set content. It shows that the dataset contains 8 variables with 96 participants recruited. A further analysis shows that of these observations, 185 are missing values corresponding to 37 participants.These observations with missing values are discarded leaving the dataset with 59 observations.

```{r}
df <- na.omit(df)
head(df)

str(df)

writeLines(paste("\nMissing values:", sum(is.na(df))))
```

# Question I: Effect of sex on flow at T0

The differences of blood flow at T0 due to sex will be analyzed using two-tailed unpaired t-test and a one-way ANOVA. Normality of the data is assumed and statistical significance will be interpreted at alpha = 0.05. The results of the two statistical tests will be compared and discussed.

## a) t-test

```{r}
writeLines("--------------------Male--------------------")
writeLines(paste("Mean:", mean(df[df$Sex == "M",]$T0)))
writeLines(paste("Standard Deviation:", sd(df[df$Sex == "M",]$T0)))

writeLines("\n-------------------Female-------------------")
writeLines(paste("Mean:", mean(df[df$Sex == "F",]$T0)))
writeLines(paste("Standard Deviation:", sd(df[df$Sex == "F",]$T0)))

writeLines("\n-------------------t-test-------------------")


#t test
t.test(T0~Sex,
       data = df,
       var.equal = TRUE,
       paired = FALSE)
```

After the t-test, the blood flow in males was found to be higher compared to female counter parts at T0 (133.02 ± 42.05 vs 86.71 ± 40.45; t = -4.148, p < 0.01).

## b) One-Way ANOVA

```{r, warning = FALSE}
df1.ez <- ezANOVA(data = df,
                 dv = T0,
                 wid = ID,
                 between = Sex,
                 type = 3,
                 return_aov = TRUE)
print(df1.ez)
```
A significant main effect for Sex was found in the dataset (F=17.21; P<0.01).

The levene's test for homogeneity of variance reveals that the variances are not unequal (p = 0.16106).

No post-hoc test is further needed as the ANOVA was performed using only two groups implying that the differences observed are between these two groups. 

After the one-way ANOVA test, the blood flow of males was found to be higher compared to female counter parts (133.02 ± 42.05 vs 86.71 ± 40.45; F = 17.21, p < 0.01).

## c) Comparison between t-test and ANOVA results

The t value of the t-test is -4.148 and the f-value of the ANOVA is 17.207 which happens to be t^2 and both results have the same p-value. This signifies that in the special case of two groups the F statistic reduces to be the same as the square of the t statistic. 

# 2) Does Flow differ by exercise condition?

The differences in blood flow due to exercise condition will be analyzed using a one-way ANOVA. The Post-Hoc analysis will be performed using t-tests with Bonferroni correction and using Tukey’s Honestly Significant Difference (HSD) method. The results of the two methods will be compared and discussed. Statistical significance will be interpreted at alpha = 0.05. 

## a)	One-way ANOVA

```{r 2) ANOVA, warning = FALSE}
df1.long <- gather(df, time, flow, T0:T120, factor_key = TRUE)

df1.ez <- ezANOVA(data = df1.long,
                 dv = flow,
                 wid = ID,
                 between = Cond,
                 type = 3,
                 return_aov = TRUE)
print(df1.ez)
```

A significant main effect for Exercise condition was found in the dataset (F=2.822; P<0.05).

To be more confident about these results, the assumptions regarding the homogeneity of the Variance (HoV) and the normality of the residuals have to be tested.

The levene's test for homogeneity of variance reveals that the variances are not unequal (p = 0.16106).

Normality is tested by observing the q-q plot, analyzing the skewness and kurtosis and by applying the Kolmogorov-Smirnov (KS) test.

```{r}
#Check normality of residuals
qqPlot(df1.ez[["aov"]][["residuals"]])

writeLines(paste("Skewwness:", skewness(df1.ez[["aov"]][["residuals"]])))

writeLines(paste("Kurtosis:", kurtosis(df1.ez[["aov"]][["residuals"]])))


writeLines("--------------------Test--------------------")
ks.test(x = df1.ez[["aov"]][["residuals"]],
        y = "pnorm",
        mean = mean(df1.ez[["aov"]][["residuals"]]),
        sd = sd(df1.ez[["aov"]][["residuals"]]))
```

The skewness and kurtosis are less than 1, indicating that it is possible that the distribution is normal. The q-q plot also depicts a normal distribution. Finally, the KS test indicates that the null hypothesis cannot be rejected (D=0.05, p=0.99), and thus can be said that the data follows a normal distribution.

## b)	Post Hoc analysis using t-tests with Bonferonni correction

```{r}
df1.cond <- df1.long %>% group_by(Cond) %>% summarise(mean = mean(flow), std = sd(flow))
df1.cond
```

```{r}
pairwise.t.test(x = df1.long$flow,
                g = df1.long$Cond,
                paired = FALSE,
                alternative = "two.sided",
                p.adjust.method = "bonferroni") #or "holm"
```

Post-hoc analysis using t-tests with Bonferroni correction revealed that the blood flow for weights was significantly larger than for bike (117.70 ± 50.25 vs 96.17 ± 36.04; P<0.01), for rest (117.70 ± 50.25 vs 86.37 ± 32.47; P<0.01) and for sprint (117.70 ± 50.25 vs 94.97 ± 40.33; P<0.01).

